maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-04-09 docs/paper/paper.tex
fix section references
minor changes to implementation section

\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{calc}
\tikzset{ % Define a TikZ style for an external hyperlink node.
  hyperlink node url/.style={
    alias=sourcenode,
    append after command={
      let \p1 = (sourcenode.north west),
          \p2 = (sourcenode.south east),
          \n1 = {\x2-\x1},
          \n2 = {\y1-\y2} in
      node[inner sep=0pt, outer sep=0pt, anchor=north west, at=(\p1)]
          {\href{#1}{\XeTeXLinkBox{\phantom{\rule{\n1}{\n2}}}}}
    }
  }
}
\providecommand{\XeTeXLinkBox}[1]{#1}
\newcommand{\docslink}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{#2}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}), but provides over 5.8 million unique formats for
  converting mazes to a text stream, detailed in
  \autoref{sec:tokenized-output-formats}.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. We use a highly efficient method detailed in
  \autoref{sec:implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

This package utilizes a simple, efficient representation of mazes as
subgraphs of a finite lattice, which we call a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze}{\texttt{LatticeMaze}}.
Using an adjacency matrix would be memory inefficient by failing to
exploit the highly sparse structure -- for example, for a 2-dimsnionsal
maze, only 4 of the diagonals would be filled in. On the other hand,
using an adjacency list could lead to a poor lookup time for whether any
given connection exists.

Instead, we describe mazes with the following representation: for a
\(d\)-dimensional lattice with \(r\) rows and \(c\) columns, we
initialize a boolean array \[
  A = \{0, 1\}^{d \times r \times c}
\] which we refer to in the code as a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.connection_list}{\texttt{connection\_list}}.
In two dimensions, the value at \(A[0,i,j]\) determines whether a
\emph{downward} connection exists from node \([i,j]\) to \([i+1, j]\).
Likewise, the value at \(A[1,i,j]\) determines whether a
\emph{rightward} connection to \([i, j+1]\) exists. Thus, we avoid
duplication of data about the existence of connections, at the cost of
requiring additional care with indexing when looking for a connection
upwards or to the left. Note that this setup allows for a periodic
lattice. This format generalizes to lattices of higher dimensions, but
our package provides only limited support for generation and
visualization of higher dimensional mazes. Generation of mazes is
detailed in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMazeGenerators}{\texttt{LatticeMazeGenerators}}.

To produce solutions to mazes, two points are selected uniformly at
random without replacement from the connected component of the maze, and
the \(A^*\) algorithm (\protect\hyperlink{ref-A_star}{Hart et al.,
1968}) is applied to find the shortest path between them. The endpoint
selection can be controlled via
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}},
and complications caused by this are detailed in
\autoref{sec:success-rate-estimation}.

>>>>

\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{calc}
\tikzset{ % Define a TikZ style for an external hyperlink node.
  hyperlink node url/.style={
    alias=sourcenode,
    append after command={
      let \p1 = (sourcenode.north west),
          \p2 = (sourcenode.south east),
          \n1 = {\x2-\x1},
          \n2 = {\y1-\y2} in
      node[inner sep=0pt, outer sep=0pt, anchor=north west, at=(\p1)]
          {\href{#1}{\XeTeXLinkBox{\phantom{\rule{\n1}{\n2}}}}}
    }
  }
}
\providecommand{\XeTeXLinkBox}[1]{#1}
\newcommand{\docslink}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{#2}}
\newcommand{\docslinkcode}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{\texttt{#2}}}
\newcommand{\secref}[1]{\hyperref[#1]{section: \textit{\nameref{#1}}}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}), but provides over 5.8 million unique formats for
  converting mazes to a text stream, detailed in
  \hyperref[sec:tokenized-output-formats]{section: \textit{\nameref{sec:tokenized-output-formats}}}.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. We use a highly efficient method detailed in
  \hyperref[sec:implementation]{section: \textit{\nameref{sec:implementation}}}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

This package utilizes a simple, efficient representation of mazes as
subgraphs of a finite lattice, which we call a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze}{\texttt{LatticeMaze}}.
Using an adjacency matrix for storing mazes would be memory inefficient
by failing to exploit the highly sparse structure -- for example, for a
2-dimensional maze, only 4 of the diagonals would be filled in. On the
other hand, using an adjacency list could lead to a poor lookup time for
whether any given connection exists.

Instead, we describe mazes with the following representation: for a
\(2\)-dimensional lattice with \(r\) rows and \(c\) columns, we
initialize a boolean array \[
  A = \{0, 1\}^{2 \times r \times c}
\] which we refer to in the code as a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.connection_list}{\texttt{connection\_list}}.
The value at \(A[0,i,j]\) determines whether a \emph{downward}
connection exists from node \([i,j]\) to \([i+1, j]\). Likewise, the
value at \(A[1,i,j]\) determines whether a \emph{rightward} connection
to \([i, j+1]\) exists. Thus, we avoid duplication of data about the
existence of connections and facillitate fast lookup time, at the cost
of requiring additional care with indexing. Note that this setup allows
for a periodic lattice. Generation of mazes is detailed in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMazeGenerators}{\texttt{LatticeMazeGenerators}}.

To produce solutions to mazes, two points are selected uniformly at
random without replacement from the connected component of the maze, and
the \(A^*\) algorithm (\protect\hyperlink{ref-A_star}{Hart et al.,
1968}) is applied to find the shortest path between them. The endpoint
selection can be controlled via
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}},
and complications caused by this are detailed in
\hyperref[sec:success-rate-estimation]{section: \textit{\nameref{sec:success-rate-estimation}}}.
A maze with a solution is denoted a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}},
which inherits from
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze}{\texttt{LatticeMaze}}.
